---
title: 'Practicals for the EEC-Master module Macroecology and global change'
author: "Ecology & Macrecology group, University of Potsdam"
output: 
  html_document:
    toc: no
    toc_float: false

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r HUlogo, echo=F, out.width="15%"}
knitr::include_graphics("figures/Mathnatlogo.svg")
```



The Master module "Macroecology and global change" is an elective module within the Master programme "Ecology, Evolution and Conservation (EEC)" at the University of Potsdam. It is typically run in the winter term.

- **Module coordinator:** Prof. Dr. Damaris Zurell

This site provides the accompanying R practicals for the module. Topics include:

- **Macroecological analyses**: working with spatial data in R, species range maps, species richness gradients, species richness models
- **Species distribution modelling**: niche concept, general modelling cycle, advanced algorithms, ensemble modelling


Prerequisites for the module are: basic understanding of macroecology (e.g. from accompanying lecture), basic knowledge of statistics, basic knowledge in geographic information systems, basic understanding of R. If you don't feel comfortable with R yet, you should work through the [*R preparatory course*](https://damariszurell.github.io/EEC-R-prep/) first. 

