---
title: "Spatial data in R"
output: 
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
bibliography: mgc.bib     
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

<div class="alert alert-info">
**RStudio project**

Open the RStudio project that we created in the previous session. I recommend to use this RStudio project for the entire module and within the RStudio project create separate R scripts for each session. 

- Create a new empty R script by going to the tab "File", select "New File"  and then "R script"
- In the new R script, type `# Session a1: Spatial data in R` and save the file in your folder "scripts" within your project folder, e.g. as "a1_SpatialData.R"
</div>


Nowadays, R offers a lot of GIS functionalities in different packages. 

We will use a number of such GIS packages in R. Remember that we can install new packages with the function `install.packages()`. Make sure also to install the dependencies.

```{r eval=F}
install.packages(c('sp','raster','dismo'), dep=T)
```

# Introduction to spatial data

Spatial data are typically stored as either vector data or raster data [@With2019]. Discrete objects with clear boundaries are usually represented by vector data (e.g. individual trees, roads, countries) while continuous phenomena without clear boundaries are usually represented by raster data (e.g. elevation, temperature). 

**Vector data** can be mapped a points, lines, and polygons:

- **Points** occur at discrete locations and are represented by a coordinate pair (x,y). Each point may carry additional informaton, the so-called attributes, e.g. the species name of the individual captured at the location along with date. 
- **Lines** describe linear features and are defined by at least two coordinate pairs (x,y), the end points of the line. A line can also consistent of several line segments.
- **Polygons** describe two-dimensional features in the landscapes and define a bounded area, enclosed by lines. Thus, a polygon needs to consist of at least three coordinate pairs (x,y.

By contrast, **Raster data** represent the landscape as a regularly spaced grid and are used to represent data that vary continuously across space such as elevation, temperature or NDVI. Raster cells can contain numeric values (e.g. elevation) or categorical values (e.g. land cover type). The coordinate information is stored differently from vector data because storing the coordinates for each grid cell in the raster would use too much storage space. To define a raster, we only need to know the coordinates of one corner, the spatial extent and the spatial resolution to infer the coordinates of each cell. 



# Vector data in R

The package `sp` defines a set of classes to represent spatial vector data and perform spatial data analyses. In this package, the geometries of vector data are stored as `SpatialPoints`, `SpatialLines` and `SpatialPolygons`, and their attributes are stored within `SpatialPointsDataFrame`, `SpatialLinesDataFrame` and `SpatialPolgyonsDataFrame`

For more detailed tutorials see https://rspatial.org. 

## SpatialPoints

For illustration, we will generate some random points data and convert them into `Spatial*` objects.

```{r}
# We set a seed for the random number generator, so that we all obtain the same results
set.seed(12345)

coords <- cbind(
  x <- rnorm(10, sd=2),
  y <- rnorm(10, sd=2)
)

str(coords)
plot(coords)
```

We now convert the random points into a `SpatialPoints` object and inspect it.
```{r}
library(sp)

# Convert into SpatialPoints
sp <- SpatialPoints(coords)

# Check out the object
class(sp)

# Inspect the (spatial) information in the object:
showDefault(sp)

# The raster package also provides a nicer print summary, so we load it as well:
library(raster)
sp
```

The slot `bbox` describes the bounding box around the known coordinates and the slot `proj4string` stores the coordinate reference system ("CRS"). In our case, no CRS was provided. We can explicitly add a CRS when creating the `SpatialPoints` object:

```{r}
sp <- SpatialPoints(coords, proj4string=CRS('+proj=longlat +datum=WGS84'))
sp
```

Next, we can generate a `SpatialPointsDataFrame` containing additional information (or attributes) for each point. Let's assume the spatial points are trees, either of three species. 

```{r}
# Create attribute table
(data <- data.frame(ID=1:10,species=sample(c('beech','oak','birch'),10,replace=T)))

# Create SpatialPointsDataFrame
(spdf <- SpatialPointsDataFrame(sp, data))

# To see what is inside
str(spdf)
```

## SpatialLines and SpatialPolygons

The structure of `SpatialLines` and `SpatialPolygons` objects is analogous to `SpatialPoints`. We use the functions `spLines()` and `spPolygons()` from the `raster` package to creates these objects.

```{r warning=F}
# Create SpatialLines through all oak trees in the data
lns <- spLines(subset(spdf,species=='oak'), crs=CRS('+proj=longlat +datum=WGS84'))


# Create SpatialPolygons for all birch trees in the data
pols <- spPolygons(subset(spdf,species=='birch'), crs=CRS('+proj=longlat +datum=WGS84'))
```

Finally, let's plot our point, line and polygon data.
```{r}
plot(sp, axes=T, lwd=2)
plot(lns, add=T, col='red', lwd=2)
plot(pols, border='blue',col='yellow',add=T)
```

## Reading vector data from file
Most often, vector data are stored as shapefiles. Download the [zip folder](data/IUCN_Sorex_alpinus.zip) containg IUCN range data of the Alpine Shrew to your data folder and unzip it. You will see that several files are contained in this folder. All of these are necessary parts of the shapefile and contain the different information on geometry and attributes. 

We use the `raster` package to read in the data:
```{r}
(shrew <- shapefile('data/IUCN_Sorex_alpinus.shp'))

# Plot Central Europe
library(maps)
map('world',xlim=c(5,30), ylim=c(40,55))

# Overlay the range of the Alpine Shrew
plot(shrew, col='red', add=T)

```

# Raster data in R
We will use the `raster` package to represent and analyse raster data in R. The package contains different raster data classes, most important for this course are `RasterLayer`, `RasterStack` and `RasterBrick`. `RasterLayer` contain only a single layer of raster information while `RasterStack` and `RasterBrick` can contain multiple layers (from separate files or from a single multi-layer file, respectively).

The function `raster()` can be used to create or read in `RasterLayer` objects. 

```{r}
(r1 <- raster(ncol=10, nrow=10, xmx=-80, xmn=-150, ymn=20, ymx=60))
```

We can access the attributes of each grid cell by using the function `values()`. Obviously, there are no values yet in the `RasterLayer` object and thus we assign some randomly.

```{r}
summary(values(r1))
values(r1) <- rnorm(ncell(r1))

# plot the raster
plot(r1)
```

A `RasterStack` can be created or read in using the function `stack()`.

```{r}
# Create another RasterLayer and assign values
r2 <- r1
values(r2) <- 1:ncell(r2)

# Stack the raster layers
(r <- stack(r1,r2))
plot(r)
```

## Read in raster data
In most cases, we will read in raster data from file. For this we can use the same commands as above, `raster()` for reading in single-layer raster files and `stack()` or `brick()` for reading in multi-layer raster files.

First, download the temperature map for UK [here](data/UK_temp.tif) to your data folder and read it in. The raster layer represents the mean annual in temperature [°C * 10]. 

```{r}
(temp <- raster('data/UK_temp.tif'))

plot(temp)
```

Second, download the [zip folder](data/UK_bioclim.zip) containing multi-layer raster data for 19 bioclimatic variables of UK, and unzip in your data folder. It contains two files. The file ending "\*.grd" represents the natural file ending for files created using the `raster` package. It needs to be accompanied by a file ending on "\*.gri", which contains all the cell attributes. The bioclimatic variables are explained [here]()https://www.worldclim.org/data/bioclim.html.

```{r}
(bioclim <- stack('data/UK_bioclim.grd'))

plot(bioclim)
```

## Download raster data
The `raster` package is offering direct access to some standard repositories using the function `getData()`, for example altitudinal data, data on the global administrative boundaries as well as current and future climates from [WorldClim](http://worldclim.org/) [@Hijmans2005]. For more information, see the help pages `?getData`. 

```{r echo=F}
clim <- getData("worldclim", var="bio", res=10, download=F, path="/Users/zurell/data/Lehre/HU_Bsc_Msc/HU_M8_GCIB/GCIB_2019/data/clim_data")
clim
```

```{r eval=F}
# Download global bioclimatic data from worldclim:
(clim <- getData("worldclim", var="bio", res=10, download=T, path="data"))
```
```{r}
plot(clim)
```

## Manipulate raster data
The `raster` package offers different functionalities to manipulate the spatial data, for example aggregating the data to coarser resolutions (`aggregate()`), or cropping to a specific extent (`crop()`).

```{r}
# Crop the temperature layer (bio1) to roughly European extent
temp_eur <- crop(clim[[1]], c(-15,45,35,72))

# Aggregate to one-degree and two-degree resolution
temp_eur_onedeg <- aggregate(temp_eur, 6)
temp_eur_twodeg <- aggregate(temp_eur, 12)

par(mfrow=c(1,3))
plot(temp_eur)
plot(temp_eur_onedeg)
plot(temp_eur_twodeg)
```

## Extract raster data
There are different ways for extracting information from raster layers. We have already worked with `values()`. If we have coordinate data, we can use these coordinates to “pierce” through raster layers. That’s one of the easiest ways to extract relevant environmental data for specific locations. For illustration, we create some random locations and then extract the climate data from our raster layers.

```{r}
# Generate random locations
library(dismo)
lonlat <- randomPoints(temp_eur,10)

# Map temperature and locations
plot(temp_eur)
points(lonlat, cex=2, pch=19)

# Extract temperature values at these locations
raster::extract(temp_eur, lonlat)
```

Note: if no environmental data are available for a specific location, e.g. no temperature values are available in the ocean, then these locations will receive an `NA` value. Hence, you should always check your resulting data for `NA`s. 

<div class="alert alert-info">
_**Exercise:**_

- Take a look at the description of [Coordinate Reference Systems, CRS, by Robert Hijmans](https://rspatial.org/raster/spatial/6-crs.html#)
- What are CRS? How can I extract the CRS information from a raster object?
- Which CRS is used in the `temp_eur` raster?
- Reproject `temp_eur` to a different CRS following instructions on above-mentioned website
- When is reprojection useful?
</div>


# References