---
title: "Species range maps"
output: 
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
bibliography: mgc.bib     
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

<div class="alert alert-info">
**RStudio project**

Open the RStudio project that we created in the first session. I recommend to use this RStudio project for the entire course and within the RStudio project create separate R scripts for each session. 

- Create a new empty R script by going to the tab "File", select "New File"  and then "R script"
- In the new R script, type `# Session a4: Species range maps` and save the file in your folder "scripts" within your project folder, e.g. as "a4_RangeMaps.R"
</div>


We are living in an age of big data where biodiversity are becoming increasingly available in digital format and at a global scale [@W_est_2020]. Many different types of biodiversity data exist, e.g. from standardised monitoring schemes, citizen science platforms, or expert knowledge. Each of these data types comes with own challenges. Within this module, I want to provide some first impressions on how you can obtain and process typical types of species data. Specifically, in this session, we will work with range maps of terrestrial animals and plants, while  occurrence data for terrestrial species in [GBIF](https://www.gbif.org/) (the Global Biodiversity Information facility) and for marine species in [OBIS](https://obis.org/) (the Ocean Biodiversity Information System) will be covered later in the [*species distribution modelling* part](b1_SpeciesData.html). 


# Obtaining range maps

We rarely have detailed biodiversity data available over large geographic extents. At broad (continental to global) extents, expert-drawn range maps (also called extent-of-occurrence maps) are often the primary data source on species distributions. 

## IUCN range maps

The IUCN (the International Union for the Conservation of Nature) provides expert range maps for a large number of species including mammals, birds (through [BirdLife International](http://www.birdlife.org/)), amphibians, reptiles, and freshwater fishes. There are also some range maps on plants and marine species, but these are very limited taxonomically. Have a look for which taxa range maps are available: https://www.iucnredlist.org/resources/spatial-data-download.
You can download them for free, but you should provide some information on your work to obtain the data. 

Most of the IUCN data are provided in the form of shapefiles. In practical 2, we have already loaded the range map of the [Alpine shrew](https://damariszurell.github.io/EEC-Macro/data/IUCN_Sorex_alpinus.zip) and learned how to use the package `terra` for reading in the shapefiles. You may remember that the shapefile is recognized as `SpatVector` object. 

```{r message=F}
library(terra)

# Load the shapefile
(shrew <- terra::vect('data/IUCN_Sorex_alpinus.shp'))

# Plot the Central Europe
library(maps)
map('world',xlim=c(5,30), ylim=c(40,55))

# Overlay the range of the Alpine Shrew
plot(shrew, col='red', add=T)
```

Unfortunately, there is no API for the IUCN range maps. So, you need to register with IUCN and then you can download range maps for different taxonomic groups. The range maps for birds come as separate shape files per species, while the range maps of mammals are provided as single shapefile that contains all species. Course participants can download the mammal shapefile from the moodle course - of course, the license agreement by the IUCN apply!


```{r echo=F}
path <- '/Users/zurell/data/Projects/database/IUCN/mammals_terr/'
mammals <- terra::vect(paste0(path,'MAMMTERR.shp'))
```

```{r eval=F}
# Read shapefile for all mammals using the raster package:
mammals <- terra::vect('data/MAMMTERR.shp')
```

The shapefile contains the range polygons for all described mammal species. The attribute table contains information on species' PRESENCE, ORIGIN, and SEASONALity. Please have a look at the metadata to understand the different values these attributes can be coded as. 

```{r}
mammals

# Inspect attribute table
terra::head(mammals)
```

We can search for specific species or species groups in the attribute table in different ways:

```{r eval=F}
# Range map for the species 'Lynx lynx'
terra::subset(mammals, mammals$BINOMIAL == "Lynx lynx")

# Show all entries for the species with the word 'Lynx' in their name
grep('Lynx',mammals$BINOMIAL, value=T)

# Range map for all species with the word 'Lynx' in their name
mammals[grep('Lynx', mammals$BINOMIAL),]
```

We can use the attribute table subsets to select specific polygons that we want to look at.

```{r}
# Assign range map of the Eurasian lynx to separate object
lynx_lynx <- mammals[mammals$BINOMIAL=='Lynx lynx',]

# Map the range
map('world')
plot(lynx_lynx, col='red', add=T)
```

<div class="alert alert-success" role="alert">
  **Test it yourself**

- Select another mammal species and plot the range map.
</div>


## BIEN range maps

The [BIEN](https://bien.nceas.ucsb.edu/bien/biendata/) database (Botanical Information and Ecology Network) contains many range maps on plants, but unfortunately only for the Americas. These can be accessed using the `BIEN` package. As illustrative example, we load the range map for the monkey puzzle tree (or Chilean pine - *Araucaria araucana*).

```{r message=F}
library(BIEN)

# Load the range map for the monkey puzzle
(monkey_puzzle_raster <- BIEN_ranges_load_species('Araucaria_araucana'))
```

Currently, the `BIEN`package still relies on the spatial R package `raster` that was the predeccesor of `terra` and is still being used by many packages. Here, we simply convert the `SpatialPolygonsDataFrame` object into the `SpatVector`object that we already know from the `terra` package.

```{r warning=F}
# Change into SpatVector object (terra package)
(monkey_puzzle_terra <- terra::vect(monkey_puzzle_raster, crs='+proj=longlat +datum=WGS84'))

# Map
map('world',xlim = c(-180,-20),ylim = c(-80,80))
plot(monkey_puzzle_terra,col='red',add=T)
```

The native range of the mokey puzzle is in the Chilean Andes, so clearly the range maps also show areas where the species naturalized.

# Working with range maps

## Range size and range centre

The `terra` package allows us to easily calculate the area of the polygons, meaning the range size of our species. The function `expanse()` outputs the area in square meters, kilometers or hectars. 

```{r}
# Range area of alpine shrew in square kilometers:
terra::expanse(shrew, unit="km")

# Range area of monkey puzzle in square kilometers:
terra::expanse(monkey_puzzle_terra, unit="km")
```

We can also very easily calculate the centre of gravity or range centroid from the `SpatVector` object. 

```{r}
# Range centroid:
terra::centroids(shrew)

# Map the species range and add the centroid to the map
map('world',xlim=c(5,30), ylim=c(40,55))
plot(shrew, col='red', add=T)
points(terra::centroids(shrew), col='blue',cex=3,pch="+")
```

We need to be careful how to interpret these centroids. They represent the centre of gravity, i.e. the mean coordinates of the distribution (weighted by cell size) but obviously, if we have several patches, the centroid might not even fall within an occupied patch. 

In case of the Monkey Puzzle, we have suspected that the range maps also contain non-native areas. We can clip the range maps to a desired spatial extent and then only calculate the range centroid for this (presumed) native range. 

```{r message=F, warning=F}
# Let's crop the range polygons to (rough coordinates of) South America
monkey_puzzle_SAm <- terra::intersect(monkey_puzzle_terra, ext(-85, -30, -55, 5))

# Map the range and range centroid
map('world',xlim = c(-100,-10),ylim = c(-60,15))
plot(monkey_puzzle_SAm,col='red',add=T)
points(terra::centroids(monkey_puzzle_SAm), col='blue',pch='+', cex=3)
```

<div class="alert alert-success" role="alert">
  **Test it yourself**

- Use the range map of the mammal species you selected earlier, and calculate the range size and range centroid. Add the point location of your centroid to the range map that you have plotted previously.
</div>


## Rasterising range maps
For many applications in macroecology, we need to rasterise the polygons. The problem is that it is unclear at which spatial resolution the range maps accurately represent species occurrences. @Hurlbert2007 and @Jetz2012a define the minimum spatial resolution as 100-200km (1-2°), although also resolutions of 50km (0.5°) and finer have been used [@Krosby2015; @Zurell2018c]. 

### Rasterising range maps with `terra`
Rasterising polygon data is made very easy in the `terra` package. We first have to define a `SpatRaster` object of the desired resolution, and then transfer the polygon data to the raster cells. 

```{r}
# By default, terra() will create a 1° resolution map in the *WGS 84* coordinate system (lon/lat).
(r_1deg <- terra::rast())

# Now, rasterise the shrew polygon data to the 1° raster grid
(shrew_1deg <- terra::rasterize(shrew, r_1deg))

map('world',xlim=c(5,30), ylim=c(40,55))
plot(shrew, col='red', add=T)
plot(shrew_1deg, add=T, alpha=0.6, legend=F)
```

Obviously, the margins of the range polgyon and the raster map differ at several places. 

<div class="alert alert-success" role="alert">
  **Test it yourself**

Check out the help page `?rasterize` and find out what the argument `touches` is doing. Rasterise the shrew range map again with setting a different `touches` argument and map the result. What is the difference?
</div>

We look at a second example, the lynx:

```{r}
# Define an empty SpatRaster of the world at 2° spatial resolution
(r_2deg <- terra::rast(res=2))

# Rasterize the eurasian lynx data
lynx_lynx_2deg <- terra::rasterize(lynx_lynx, r_2deg)

# Map the occupied grid cells
map('world')
plot(lynx_lynx_2deg, add = TRUE, legend = FALSE)
```


### Rasterising range maps with `letsR`

There are also specific macroecological packages in R that facilitate working with range maps and rasterising them, for example the function `lets.presab()` in the `letsR` package. At the moment, this package still relies on the older spatial R package `raster`.

```{r warning=F}
library(letsR)
library(raster)

# The lets.presab() function expects "binomial" as one specific column name in the Polygons data frame
values(shrew)
values(monkey_puzzle_terra)
monkey_puzzle_raster@data

# As the BIEN data do not contain "binomial" as column name, we have to add it:
colnames(monkey_puzzle_raster@data) <- "binomial"

# We set the resolution to 1 degree (the default) and restrict the spatial extent to South America
r_monkey_puzzle <- lets.presab(monkey_puzzle_raster, resol=1, xmn = -100, xmx = -10, ymn = -57, ymx = 15)

# This time, we receive a raster object (instead of SpatRaster)
r_monkey_puzzle

# Map the range and range centroid
map('world',xlim = c(-100,-10),ylim = c(-60,15))
plot(monkey_puzzle_SAm,col='blue',add=T)
plot(r_monkey_puzzle, add=T, alpha=0.6, legend=F)
```


<div class="alert alert-success" role="alert">
  **Test it yourself**

- Rasterise the range map of the mammal species you selected above. Try both approaches using the `terra` package and the `letsR` package.
</div>


### Bulk-rasterising multiple species range maps with `letsR`

The `letsR` package also allows to bulk-download multiple species and rasterise them to form a richness map. As first example, we look at the *Pinus* genus in the Americas

```{r warning=F}
# Extract the available Pinus species names
(pinus_names <- BIEN_ranges_genus("Pinus",match_names_only = T)[,1])

# Download the range maps for all Pinus species (as SpatialPolygonsDataFrame using the raster standard instead of terra)
(pinus <- BIEN_ranges_load_species(pinus_names))

# As the BIEN data do not contain "binomial" as column name, we have to re-format the column names again before rasterising
colnames(pinus@data) <- "binomial"
r_pinus <- lets.presab(pinus, resol=1, xmn = -170, xmx = -10, ymn = -57, ymx = 60)

# Plot species richness
plot(r_pinus)
```

In the `lets.presab()` function, we can also specify which PRESENCE, ORIGIN and SEASONAL information should be used and which not (see IUCN metadata on moodle). Let's look at Neotropical fruit bats (**Artibeus**) as example. Here, we set `presence=1` meaning that we only consider extant species, `origin=1` meaning only native species, and `seasonal=1` meaning only resident species.

As the `letsR` package does not work with `SpatRaster` objects yet, we have to use the older `raster` package to prepare the range map data.

<!-- ```{r echo=F} -->
<!-- (mammals_raster <- raster::shapefile(paste0(path,'MAMMTERR.shp'))) -->
<!-- ``` -->
<!-- ```{r echo=F, eval=F} -->
<!-- library(raster) -->
<!-- (mammals_raster <- as(mammals,"Spatial")) -->
<!-- # first subset SpatVector and then convert to SpatialPolygonsDataFrame -->
<!-- artibeus_spp <- as(mammals[grep('Artibeus',mammals$BINOMIAL),], "Spatial") -->
<!-- ``` -->
<!-- ```{r eval=F, echo=F} -->
<!-- # prepare raster version of the mammals shapefile -->
<!-- (mammals_raster <- raster::shapefile('data/MAMMTERR.shp')) -->
<!-- ``` -->

```{r warning=F}
library(raster)

# Subset the SpatVector
artibeus_spp <- mammals[grep('Artibeus',mammals$BINOMIAL),]

# Convert to SpatialPolygonsDataFrame from raster package
artibeus_spp <- as(artibeus_spp, "Spatial")

# Rasterize the ranges using the letsR package
r_artibeus_spp <- lets.presab(artibeus_spp, resol=2, 
                              presence = 1, origin = 1, seasonal = 1)

# Map the species richness
plot(r_artibeus_spp)

# Map single species - here, just the first two
par(mfrow=c(1,2))
plot(r_artibeus_spp, name = "Artibeus amplus")
plot(r_artibeus_spp, name = "Artibeus anderseni")

# Look at structure of the object and at the presence-absence matrix
str(r_artibeus_spp, 1)
head(r_artibeus_spp$Presence_and_Absence_Matrix)
```


<div class="alert alert-info">
_**Exercise:**_

- Select another genus of plants or mammals, and follow the workflow to rasterise range maps (either using `terra` or `letsR`) and map species richness of the species within that genus.
- Plot the latitudinal species richness gradient for this genus. Interpret in light of previously discussed patterns.
</div>


# References